Brownian motion of main-belt asteroids on human timescales

### 2025-03-31 main.tex
Fix Yarkovsky discussion

\documentclass[onecolumn]{aastex631}
%\documentclass[linenumbers, onecolumn]{aastex631}

@@

\begin{abstract}
  The explosion of high-precision astrometric data on main-belt
  asteroids (MBAs) enables new
  inferences on gravitational and non-gravitational forces present in
  this region.  We estimate the size of MBA motions
  caused by mutual gravitational encounters with other MBAs that are
  either omitted from ephemeris models or have uncertain mass
  estimates. In other words, what is the typical Brownian motion among
  MBAs that cannot be predicted from the ephemeris, and therefore
  serves as noise on inferences from the MBAs?  We estimate the RMS
  azimuthal shift $\sigma_\phi$ of this ``Brownian noise''  by
  numerical estimation of the distribution of impulse sizes and
  directions among known MBAs, combined with analytical propagation
  into future positional uncertainties. At current
  levels of asteroid-mass knowledge, $\sigma_\phi$ rises to
  $\approx2$~km or $\approx1$~mas over $T=10$~yr, increasing as
  $T^{3/2},$  large enough to degrade many inferences from
  Gaia and LSST MBA data. LSST data will, however, improve MBA
  mass knowledge enough to lower this Brownian uncertainty by
  $\ge4\times.$  Radial and vertical Brownian noise  at $T=10$~yr are
  factors $\approx7$ and $\approx45,$ respectively,
  lower than the azimuthal noise, and grow as $T^{3/2}$
  and $T^{1/2}.$  For full exploitation of Gaia and
  LSST MBA data, ephemeris models should include the
  $\approx1000$~largest asteroids as active bodies with free masses,
  even if not all are well constrained.  This will correctly propagate
  the uncertainties from these 1000 sources' deflections into desired
  inferences. The RMS value of deflections from less-massive MBAs is
  then just $\sigma_\phi\approx60$~m or 30~$\mu$as, small enough to ignore until occultation-based position data become ubiquitously available for MBAs.
\end{abstract}

@@

\subsection{Comparison to radiation forces}
As we consider smaller MBAs as test particles, radiation pressure becomes larger relative to gravitational forces.  It is interesting to compare the effects of radiation pressure to those of Brownian motion to see when the former are a bigger source of noise than the latter, or contrarily to see whether Brownian motion interferes with attempts to measure radiation forces on MBAs.  At first order, radiation pressure from the Sun and the reaction from reflected sunlight cause a constant radial force on the MBA.  The ratio of this force to the gravitational force for an asteroid of radius $R,$ geometric albedo $q$, and density $\rho$ is $(1+q)\Psi,$ where
\begin{equation}
 \Psi \equiv \frac{F_{\rm rad}}{F_{\rm grav}} = \frac{\pi R^2 L_\odot/4\pi c}{4\pi GM_\odot R^3\rho/3}
   = \frac{3}{16\pi} \frac{L_\odot}{GM_\odot c R \rho} \approx
   3\times10^{-10} \times 10^{0.2(H-16)},
\end{equation}
where the last expression adopts our standard values $q=0.25,
\rho=2500\,{\rm kg}\, {\rm m}^{-3}.$ The primary observational consequence of
this constant radial force is a time-independent, $H$-dependent 
shift $\Delta r = -(1+q)\Psi a$ of the radial position a body should
have at a given period.  
This does not resemble
any expected effect of Brownian motion, and also very difficult to
measure since radar ranging to MBAs is infeasible.  More important for
observations is the
Yarkovsky effect, in which anisotropic re-emission of the absorbed
radiation leads to a net force that is a fraction $g_Y(1-q)$ of the incident radiation pressure.  The Yarkovsky effect has been detected for several NEAs but no individual MBAs, with values $g_Y \approx 0.1$ \citep[Yarkovsky effect and data are review by][]{yarkovsky}.  This is also consistent with detections of the Yarkovsky effect by examining the $\Delta a$ dependence on $H$ of members of asteroid \emph{families} in the Main Belt \citep[\eg][]{karin}.

@@

Taking a case where the Yarkovsky force is in the prograde azimuthal direction, and using our system of units with $a=1$ and $v_c=1,$ this leads to a continuous power $dE/dt\approx g_Y\Psi$ which in turn implies $da/dt\approx2g_Y\Psi.$   The mean motion rate is hence $1-3 g_Y\Psi t$ and the accumulated shift in mean anomaly over time $T$ amounts to $\Delta\phi = -3 g_Y \Psi  T^2/2.$   In this case, a rough comparison of the accumulated $\Delta\phi$ from the Yarkovsky effect to the $\sigma_\phi$ of noise from unmodelled Brownian motion, after a time interval $T$, is
\begin{equation}
\frac{\textrm{Yarkovsky}}{\textrm{Brownian}} = 
   \frac{\Delta\phi}{\sigma_\phi} \approx \frac{ 3 g_Y \Psi
     T^2/2}{\sqrt{3 T^3 \nIphi}} \approx \sqrt{T/10\,{\rm yr}} \times 10^{0.2(H-20.2)}
     \label{eq:yarkovsky}
\end{equation}
if we assume $\nIphi$ from the current knowledge of asteroid masses.   There is thus a break point near $H=20$, or $\approx100$~m diameter: for MBAs larger than this, the ephemeris uncertainties from the Yarkovsky effect will be smaller than those from mutual gravitational encounters on decade timescales. This includes all MBAs detected by Gaia and nearly all detectable by LSST.  Below this size, the Yarkovsky displacement grows larger in a decade and has a better chance of being detectable in the presence of Brownian motion.  Distinguishing the steady quadratic growth of the Yarkovsky $\Delta\phi$ from the random-walk growth of Brownian motion to the same final size requires observational S/N and time resolution significantly better than is needed for detecting either one individually.

When Yarkovsky forces on MBAs are estimated using the $H$-dependent
dispersion of $a$ for asteroid families, the effective ``observation''
time period is typically several Myr.  Over these long time scales,
\eqq{eq:yarkovsky} shows that the temporally coherent Yarkovsky effect
can dominate over Brownian motion for bodies up to 10's or 100's of km
in diameter, which is why the technique is successful.

@@

Making inferences from MBA astrometry becomes more complicated when
Brownian motion is a significant contributor to the error budget, but
a good path forward is to do all model fitting by including the
$\approx1000$ largest MBAs as active bodies in the ephemeris model and
leaving their masses as free parameters.  This will constrain the
masses of these bodies to the full extent that the data allow, and
will properly propagate any remaining uncertainty in those masses into
the other inferences that one is trying to make.  The noise
contributed by the remaining millions of unmodelled MBAs would be just
$\approx60$~m RMS in the azimuthal direction (about 30~$\mu$as), and a
few $\mu$as or less in the other axes, small enough to ignore entirely
for the accuracy of imaging astrometry for the foreseeable future.
Occultation-based astrometry is much more accurate (but has fewer possible measurements per tracer), and would be significantly impacted by Brownian motion until $\gtrsim1000$ MBAs have mass estimates at modest $S/N$ level, although the precise impact will depend on whether the signals one is seeking are parallel or orthogonal to the kind of variations caused by Brownian motion.  For objects $\gtrsim100$~m in diameter, uncertainties from mutual encounters are larger than uncertainties from the Yarkovsky effect over decade timescales, at present levels of knowledge of asteroid masses.

>>>>

%\documentclass[onecolumn]{aastex631}
\documentclass[linenumbers, onecolumn]{aastex631}

@@

\begin{abstract}
  % The explosion of high-precision astrometric data on main-belt
  % asteroids (MBAs) enables new
  % inferences on gravitational and non-gravitational forces present in
  % this region.  We estimate the size of MBA motions
  % caused by mutual gravitational encounters with other MBAs that are
  % either omitted from ephemeris models or have uncertain mass
  % estimates. In other words, what is the typical Brownian motion among
  % MBAs that cannot be predicted from the ephemeris, and therefore
  % serves as noise on inferences from the MBAs?  We estimate the RMS
  % azimuthal shift $\sigma_\phi$ of this ``Brownian noise''  by
  % numerical estimation of the distribution of impulse sizes and
  % directions among known MBAs, combined with analytical propagation
  % into future positional uncertainties. At current
  % levels of asteroid-mass knowledge, $\sigma_\phi$ rises to
  % $\approx2$~km or $\approx1$~mas over $T=10$~yr, increasing as
  % $T^{3/2},$  large enough to degrade many inferences from
  % Gaia and LSST MBA data. LSST data will, however, improve MBA
  % mass knowledge enough to lower this Brownian uncertainty by
  % $\ge4\times.$  Radial and vertical Brownian noise  at $T=10$~yr are
  % factors $\approx7$ and $\approx45,$ respectively,
  % lower than the azimuthal noise, and grow as $T^{3/2}$
  % and $T^{1/2}.$  For full exploitation of Gaia and
  % LSST MBA data, ephemeris models should include the
  % $\approx1000$~largest asteroids as active bodies with free masses,
  % even if not all are well constrained.  This will correctly propagate
  % the uncertainties from these 1000 sources' deflections into desired
  % inferences. The RMS value of deflections from less-massive MBAs is
  % then just $\sigma_\phi\approx60$~m or 30~$\mu$as, small enough to
  % ignore until occultation-based position data become ubiquitously
  % available for MBAs.
%
High-precision astrometric data on main-belt
asteroids (MBAs) enables new
inferences on gravitational and non-gravitational forces present in
this region.  We estimate the typical Brownian motion due to mutual encounters among
MBAs that cannot be predicted from the ephemeris of larger bodies, and therefore
serves as noise on inferences from the MBAs.  The RMS
azimuthal shift $\sigma_\phi$ of this ``Brownian noise''  is predicted
using the distribution of known MBA orbits, combined with analytical propagation
into future positional uncertainties. At current
levels of asteroid-mass knowledge, $\sigma_\phi$ rises to
$\approx2$~km or $\approx1$~mas over $T=10$~yr, increasing as
$T^{3/2},$  large enough to degrade many inferences from
Gaia and LSST MBA data. LSST data will, however, improve MBA
mass knowledge enough to lower this Brownian uncertainty by
$\ge4\times.$  Radial (vertical) Brownian noise  at $T=10$~yr is
$\approx7\times$ ($\approx45\times$) 
lower than the azimuthal noise.  For full exploitation of Gaia and
LSST MBA data, ephemeris models should include the
$\approx1000$~largest asteroids as active bodies with free masses,
even if not all are well constrained.  This will correctly propagate
the uncertainties from these 1000 sources' deflections into desired
inferences. The RMS value of deflections from less-massive MBAs is
then just $\sigma_\phi\approx60$~m or 30~$\mu$as, small enough to ignore until occultation-based position data become ubiquitously available for MBAs.
\end{abstract}

@@

\subsection{Comparison to radiation forces}
As we consider smaller MBAs as test particles, radiation pressure becomes larger relative to gravitational forces.  It is interesting to compare the effects of radiation pressure to those of Brownian motion to see when the former are a bigger source of noise than the latter, or contrarily to see whether Brownian motion interferes with attempts to measure radiation forces on MBAs.  At first order, radiation pressure from the Sun and the reaction from reflected sunlight cause a constant radial force on the MBA.  The ratio of this force to the gravitational force for an asteroid of radius $R,$ geometric albedo $q$, and density $\rho$ is $(1+q)\Psi,$ where
\begin{equation}
 \Psi \equiv \frac{F_{\rm rad}}{F_{\rm grav}} = \frac{\pi R^2 L_\odot/4\pi c}{4\pi GM_\odot R^3\rho/3}
   = \frac{3}{16\pi} \frac{L_\odot}{GM_\odot c R \rho} \approx
   3\times10^{-10} \times 10^{0.2(H-16)},
\end{equation}
where the last expression adopts our standard values $q=0.25,
\rho=2500\,{\rm kg}\, {\rm m}^{-3}.$ The primary observational consequence of
this constant radial force is a time-independent, $H$-dependent 
shift $\Delta r = -(1+q)\Psi a$ of the radial position a body should
have at a given period.  
This does not resemble
any expected effect of Brownian motion, and also very difficult to
measure since radar ranging to MBAs is infeasible.  More important for
observations is the
Yarkovsky effect, in which anisotropic re-emission of the absorbed
radiation leads to a net force that is a fraction $g_Y(1-q)$ of the
incident radiation pressure.  The Yarkovsky effect has been detected
for several NEAs but no individual MBAs, with values $g_Y \approx 0.1$
\citep[Yarkovsky effect and data are review by][]{yarkovsky}.
\edited{A value of $g_Y\approx0.1$ is also the largest that can be obtained
  from a model of a spherical body with uniform surface properties and
  Lambertian emission.} This is also consistent with detections of the
Yarkovsky effect by examining the $\Delta a$ dependence on $H$ of
members of asteroid \emph{families} in the Main Belt
\citep[\eg][]{karin}.  \edited{We will assume an RMS value of $g_Y=0.05,$
the influence of the surface absorption coefficient being $<1.$}

@@

Taking a case where the Yarkovsky force is in the prograde azimuthal direction, and using our system of units with $a=1$ and $v_c=1,$ this leads to a continuous power $dE/dt\approx g_Y\Psi$ which in turn implies $da/dt\approx2g_Y\Psi.$   The mean motion rate is hence $1-3 g_Y\Psi t$ and the accumulated shift in mean anomaly over time $T$ amounts to $\Delta\phi = -3 g_Y \Psi  T^2/2.$   In this case, a rough comparison of the accumulated $\Delta\phi$ from the Yarkovsky effect to the $\sigma_\phi$ of noise from unmodelled Brownian motion, after a time interval $T$, is
\begin{equation}
\frac{\textrm{Yarkovsky}}{\textrm{Brownian}} = 
   \frac{\Delta\phi}{\sigma_\phi} \approx \frac{ 3 g_Y \Psi
     T^2/2}{\sqrt{3 T^3 \nIphi}} \approx \sqrt{T/10\,{\rm yr}} \times 10^{0.2(H-12.7)}
     \label{eq:yarkovsky}
\end{equation}
\edited{if we assume $\nIphi$ from the LSST-era knowledge of asteroid
  masses.   For MBAs fainter than $H\approx12.7,$ or
  $\lesssim7$~km in diameter, the RMS uncertainty in position from
  Yarkovsky forces exceeds the Brownian motion noise.   With current-day knowledge
  of asteroid masses, the threshold for Yarkovsky dominance is a bit
  fainter than $H=12.7.$  The optimist's view of this is that Brownian
  motion will not interfere with attempts to measure the Yarkovsky
  effect on individual km-scale or smaller MBAs.  The pessimists view is that uncertainty in
 the  Yarkovsky effect will limit ephemeris accuracy before Brownian
  motion will.  In either viewpoint, the steady $T^2$ accumulation of
  $\Delta\phi$ displacements from the Yarkovsky effect is
  distinguishable from the $T^{3/2}$ random walk of Brownian motion as
  data quality and duration improve, so one separate these effects to
  some extent when fitting an ephemeris---\citet{negin} quantify the
  ability to distinguish the Yarkovsky effect from mutual encounters.}

When Yarkovsky forces on MBAs are estimated using the $H$-dependent
dispersion of $a$ for asteroid families, the effective ``observation''
time period is typically several Myr.  Over these long time scales,
\eqq{eq:yarkovsky} shows that the temporally coherent Yarkovsky effect
can dominate over Brownian motion for bodies up to hundreds of km
in diameter, which is why the technique is successful.

@@

Making inferences from MBA astrometry becomes more complicated when
Brownian motion is a significant contributor to the error budget, but
a good path forward is to do all model fitting by including the
$\approx1000$ largest MBAs as active bodies in the ephemeris model and
leaving their masses as free parameters.  This will constrain the
masses of these bodies to the full extent that the data allow, and
will properly propagate any remaining uncertainty in those masses into
the other inferences that one is trying to make.  The noise
contributed by the remaining millions of unmodelled MBAs would be just
$\approx60$~m RMS in the azimuthal direction (about 30~$\mu$as), and a
few $\mu$as or less in the other axes, small enough to ignore entirely
for the accuracy of imaging astrometry for the foreseeable future.
Occultation-based astrometry is much more accurate (but has fewer
possible measurements per tracer), and would be significantly impacted
by Brownian motion until $\gtrsim1000$ MBAs have mass estimates at
modest $S/N$ level, although the precise impact will depend on whether
the signals one is seeking are parallel or orthogonal to the kind of
variations caused by Brownian motion.  \edited{For objects
  $\lesssim7$~km in diameter, uncertainties from mutual encounters are
  smaller than uncertainties from the Yarkovsky effect over decade
  timescales, at near-future levels of knowledge of asteroid masses.
  Yarkovsky forces must be included in dynamical models in order to
  reach the precision levels where Brownian motion becomes a limitation.}
